---
title: "envi: Environmental interpolation using spatial kernel density estimation"
author: 'Ian D. Buller (Github: @idblr)'
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{envi: Environmental interpolation using spatial kernel density estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = TRUE, fig.width = 6, fig.height = 6, fig.show = "hold")
```

Start with the necessary packages and seed for the vignette
```{r packages}
loadedPackages <- c("raster", "spatstat.core", "spatstat.data")
#loadedPackages <- c("envi", "raster", "spatstat.core", "spatstat.data")
invisible(lapply(loadedPackages, require, character.only = TRUE))
set.seed(1234) # for reproducibility
source('~/Documents/GitHub/envi/R/lrren.R') # REMOVE IN FINAL VIGNETTE
source('~/Documents/GitHub/envi/R/plot_obs.R') # REMOVE IN FINAL VIGNETTE
source('~/Documents/GitHub/envi/R/plot_predict.R') # REMOVE IN FINAL VIGNETTE
source('~/Documents/GitHub/envi/R/plot_cv.R') # REMOVE IN FINAL VIGNETTE
source('~/Documents/GitHub/envi/R/lrr_plot.R') # REMOVE IN FINAL VIGNETTE
source('~/Documents/GitHub/envi/R/pval_plot.R') # REMOVE IN FINAL VIGNETTE
source('~/Documents/GitHub/envi/R/lrr_raster.R') # REMOVE IN FINAL VIGNETTE
```

We use the `gorillas` data and the accompanying covariate data in `gorillas.extra` from the `spatstat.data` package [on CRAN](https://cran.r-project.org/web/packages/spatstat.data/index.html). These data are locations of nesting sites of gorillas in the Kagwene Gorilla Sanctuary in Cameroon. A detailed description and analysis of the data are reported in Funwi-Gabga and Mateu (2012) <https://doi.org/10.1007/s00477-011-0541-1>. The authors used a kernel density-based smoothing technique to detect hot-spots of nesting in the park. Here, we use another kernel density-based smoothing technique to detect hot-spots (i.e., ecological niche) of nesting within the covariate information and then predict where these hot-spots are located within the park.

We start by importing the two covariate data of class `im`:

1. Slope Angle (degrees)
2. Distance to water (meters)

```{r covariate}
slopeangle <- spatstat.data::gorillas.extra$slopeangle
waterdist <- spatstat.data::gorillas.extra$waterdist
plot(slopeangle); plot(waterdist)
```

Center and scale the covariate data.
```{r scale}
slopeangle$v <- scale(slopeangle)
waterdist$v <- scale(waterdist)
```

Convert the covariate data to class `raster`.
```{r rasterize}
slopeangle_raster <- raster(slopeangle)
waterdist_raster <- raster(waterdist)
```

Add appropriate marks to the `gorillas` data from `spatstat.data` package. These points are considered our "presence" locations. 
```{r presence}
gorillas <- unmark(spatstat.data::gorillas)
spatstat::marks(gorillas) <- data.frame("presence" = rep(1, gorillas$n),
                                        "lon" = gorillas$x,
                                        "lat" = gorillas$y)
spatstat::marks(gorillas)$slopeangle <- slopeangle[gorillas]
spatstat::marks(gorillas)$waterdist <- waterdist[gorillas]
```

Randomly draw points from the study area and add the appropriate marks. These points are considered our "absence" locations. 
```{r absence}
absence <- spatstat::rpoispp(0.00004, win = slopeangle)
spatstat::marks(absence) <- data.frame("presence" = rep(0, absence$n),
                                       "lon" = absence$x,
                                       "lat" = absence$y)
spatstat::marks(absence)$slopeangle <- slopeangle[absence]
spatstat::marks(absence)$waterdist <- waterdist[absence]
```

Combine the presence (n = `r gorillas$n`) and absence (`r absence$n`) locations into one object of class `data.frame` and reorder the features required for the `lrren` function in the `envi` package:

1. ID
2. X-coordinate
3. Y-coordinate
4. Presence (binary)
5. Covariate 1
6. Covariate 2

```{r obs_locs}
obs_locs <- spatstat::superimpose(gorillas, absence, check = FALSE)
spatstat::marks(obs_locs)$presence <- as.factor(spatstat::marks(obs_locs)$presence)
plot(obs_locs, which.marks = "presence")
obs_locs <- spatstat::marks(obs_locs)
obs_locs$id <- seq(1, nrow(obs_locs), 1)
obs_locs <- obs_locs[ , c(6, 2, 3, 1, 4, 5)]
```

Extract the prediction locations within the study area from one of the covariates.
```{r predict_locs}
predict_locs <- data.frame(raster::rasterToPoints(slopeangle_raster))
predict_locs$layer2 <- raster::extract(waterdist_raster, predict_locs[, 1:2])
```

Run the `lrren` function within the `envi` package. We use the default settings except we want to predict the ecological niche within the study area (`predict = TRUE`), we conduct n-fold cross-validation model fit diagnostics (`cv = TRUE`) by undersampling absence locations to balance the prevalence (0.5) within all testing data sets (`balance = TRUE`), and we quiet run messages (`verbose = TRUE`).
```{r lrren}
# Run lrren
test <- lrren(obs_locs = obs_locs,
              predict_locs = predict_locs,
              predict = TRUE,
              conserve = TRUE,
              cv = TRUE,
              balance = TRUE,
              verbose = FALSE)
```

We display the estimated ecological niche within a space of Covariate 1 by Covariate 2. We use the default two-tailed alpha-level (`alpha = 0.05`) and the default colors where the yellow color denotes areas with covariate data combinations where we have sparse observations. 
```{r plot_obs}
plot_obs(test)
```

We display the estimated ecological niche predicted to the study area within geographic space. We use the default two-tailed alpha-level (`alpha = 0.05`).
```{r plot_predict}
plot_predict(test)
```

We display the 10-fold cross-validation diagnostics. We use the default two-tailed alpha-level (`alpha = 0.05`).
```{r plot_cv}
plot_cv(test)
```

Based on only two covariates, slope angle and distance to water, we detected two ecological niches of gorillas compared to psuedo-absence points drawn randomly from within the park. These two ecological niches are located in many small regions of the park. The log relative risk estimate accurately predicted about 60% of the gorilla nesting sites. 

Future improvement can be made to the choice in covariates, including using the first two components of a principal components analysis of all seven covariates available in the `gorillas.extra` data in the `spatstat.data` package.
